---
title: "AI Productivity and Burnout EDA"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(scales)
library(patchwork)
library(janitor)
```

# Dataset overview
The merged dataset contains 4,500 observations and 17 variables, with no missing values.

## Load data

```{r load-data}
features <- read_csv("../data/ai_productivity_features.csv", show_col_types = FALSE) |>
  clean_names()

targets <- read_csv("../data/ai_productivity_targets.csv", show_col_types = FALSE) |>
  clean_names()

df <- features |>
  left_join(targets, by = "employee_id")

glimpse(df)
summary(df)
```

## Quick data quality checks

```{r quality-checks}
cat("Rows:", nrow(df), "\n")
cat("Columns:", ncol(df), "\n\n")

missing_summary <- df |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "column", values_to = "missing_count") |>
  arrange(desc(missing_count))

missing_summary
```


## Distribution of burnout risk score

```{r burnout-hist}
ggplot(df, aes(x = burnout_risk_score)) +
  geom_histogram(binwidth = 0.5, color = "black", fill = "sandybrown") +
  scale_x_continuous(breaks = seq(floor(min(df$burnout_risk_score, na.rm = TRUE)),
                                  ceiling(max(df$burnout_risk_score, na.rm = TRUE)),
                                  by = 0.5)) +
  labs(
    title = "Burnout Risk Score Distribution",
    x = "Burnout Risk Score",
    y = "Count"
  ) +
  theme_minimal(base_size = 12)
```

## Burnout score diagnostics

This section checks whether the spike at 10 is coming from the underlying data or from the plot.

```{r burnout-diagnostics}
df |>
  count(burnout_risk_score, sort = TRUE)

sum(df$burnout_risk_score == 10, na.rm = TRUE)

summary(df$burnout_risk_score)
```

## Distribution by job role

```{r burnout-by-role}
ggplot(df, aes(x = burnout_risk_score)) +
  geom_histogram(binwidth = 0.5, color = "black", fill = "steelblue") +
  facet_wrap(~ job_role, scales = "free_y") +
  labs(
    title = "Burnout Risk Score by Job Role",
    x = "Burnout Risk Score",
    y = "Count"
  ) +
  theme_minimal(base_size = 12)
```

## Check whether filtering changes the spike

```{r filter-check}
designer_df <- df |>
  filter(job_role == "Designer")

writer_df <- df |>
  filter(job_role == "Writer")

p1 <- ggplot(designer_df, aes(x = burnout_risk_score)) +
  geom_histogram(binwidth = 0.5, color = "black", fill = "tomato") +
  labs(title = "Designer", x = "Burnout Risk Score", y = "Count") +
  theme_minimal(base_size = 12)

p2 <- ggplot(writer_df, aes(x = burnout_risk_score)) +
  geom_histogram(binwidth = 0.5, color = "black", fill = "darkseagreen3") +
  labs(title = "Writer", x = "Burnout Risk Score", y = "Count") +
  theme_minimal(base_size = 12)

p1 + p2
```

## Productivity score distribution

```{r productivity-hist}
ggplot(df, aes(x = productivity_score)) +
  geom_histogram(binwidth = 5, color = "black", fill = "mediumpurple") +
  labs(
    title = "Productivity Score Distribution",
    x = "Productivity Score",
    y = "Count"
  ) +
  theme_minimal(base_size = 12)
```

## Relationship between burnout and productivity

```{r burnout-vs-productivity}
ggplot(df, aes(x = burnout_risk_score, y = productivity_score, color = job_role)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Burnout Risk Score vs Productivity Score",
    x = "Burnout Risk Score",
    y = "Productivity Score",
    color = "Job Role"
  ) +
  theme_minimal(base_size = 12)
```

## Correlation heatmap for numeric variables

```{r correlation-heatmap}
numeric_df <- df |>
  select(where(is.numeric))

cor_mat <- cor(numeric_df, use = "complete.obs")

cor_df <- as.data.frame(as.table(cor_mat))

ggplot(cor_df, aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient2(low = "firebrick", mid = "white", high = "navy", midpoint = 0) +
  labs(
    title = "Correlation Heatmap",
    x = NULL,
    y = NULL,
    fill = "Correlation"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Burnout by key categorical groups

```{r boxplots}
p_deadline <- ggplot(df, aes(x = deadline_pressure_level, y = burnout_risk_score, fill = deadline_pressure_level)) +
  geom_boxplot(show.legend = FALSE) +
  labs(title = "Burnout by Deadline Pressure", x = "Deadline Pressure Level", y = "Burnout Risk Score") +
  theme_minimal(base_size = 12)

p_role <- ggplot(df, aes(x = job_role, y = burnout_risk_score, fill = job_role)) +
  geom_boxplot(show.legend = FALSE) +
  labs(title = "Burnout by Job Role", x = "Job Role", y = "Burnout Risk Score") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

p_deadline + p_role
```

## Burnout against likely drivers

```{r driver-scatterplots}
p_focus <- ggplot(df, aes(x = focus_hours_per_day, y = burnout_risk_score)) +
  geom_point(alpha = 0.6, color = "steelblue4") +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(title = "Burnout vs Focus Hours", x = "Focus Hours per Day", y = "Burnout Risk Score") +
  theme_minimal(base_size = 12)

p_balance <- ggplot(df, aes(x = work_life_balance_score, y = burnout_risk_score)) +
  geom_point(alpha = 0.6, color = "darkorange3") +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(title = "Burnout vs Work-Life Balance", x = "Work-Life Balance Score", y = "Burnout Risk Score") +
  theme_minimal(base_size = 12)

p_automation <- ggplot(df, aes(x = tasks_automated_percent, y = burnout_risk_score)) +
  geom_point(alpha = 0.6, color = "darkgreen") +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(title = "Burnout vs Tasks Automated", x = "Tasks Automated (%)", y = "Burnout Risk Score") +
  theme_minimal(base_size = 12)

p_focus / p_balance / p_automation
```

## Top observations with maximum burnout

```{r max-burnout}
df |>
  filter(burnout_risk_score == max(burnout_risk_score, na.rm = TRUE)) |>
  arrange(desc(productivity_score)) |>
  select(employee_id, job_role, burnout_risk_score, productivity_score,
         deadline_pressure_level, manual_work_hours_per_week,
         learning_time_hours_per_week, focus_hours_per_day,
         work_life_balance_score)
```

## Brief interpretation notes

```{r interpretation, results='asis'}
share_at_10 <- mean(df$burnout_risk_score == 10, na.rm = TRUE)

cat(paste0(
  "- Proportion of observations with burnout risk score exactly 10: ",
  percent(share_at_10, accuracy = 0.1), "\n",
  "- The spike at 10 is coming from the data itself.\n"
))
```

### Main target distribution finding
Burnout risk is strongly right-concentrated, with 37.4% of observations at the maximum score of 10.

### Cross-group observation
This pattern persists across job roles, suggesting that high burnout is widespread rather than isolated.

### Relationship observation
Work-life balance appears negatively associated with burnout, while productivity varies across the burnout range.